---
title: "Modeling IS6812"
author: "Brian Frerichs"
date: "2025-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


****Business Problem
The Business problem is that Home Credit wants to be able to accurately predict potential new client’s abilities to repay loans. Home Credit hopes to provide a positive lending experience to the unbaked by sheltering them from predatory lending arrangements. 

****Benefit of a Solution
If they are able to minimize the amount of loans given out that the client is unable to repay, then Home Credit will minimize their own financial risk and the unbanked population will be in a better financial state.

****Success Metrics
The success metric will be the area under the ROC curve, which is a measurement of how accurate our classifier is. An AUC of 1 indicates a perfect classifier, while an AUC of .5 indicates that our model is no better than a random classifier. So in this case an AUC closer to 1 indicates that our model successfully predicts who will and will not repay loans.

****Analytics Approach
This is a supervised learning problem, and the target variable is whether the loan was repaid. This is a classification problem. 









```{r}


library(tidyverse)
library(caret)
library(pROC)
library(janitor)
library(readr)
library(dplyr)
library(caret)
library(e1071)

application_train <- read_csv("application_train.csv")


## trim data set to only include most important predictors

application_train_slim <- application_train %>%
  select(TARGET, CNT_CHILDREN, AMT_INCOME_TOTAL,AMT_CREDIT,AMT_ANNUITY,NAME_INCOME_TYPE,DAYS_BIRTH,DAYS_EMPLOYED,OCCUPATION_TYPE,CNT_FAM_MEMBERS,EXT_SOURCE_1,EXT_SOURCE_2,EXT_SOURCE_3)


## I chose to create a new column called employment duration that bins the DAYS_EMPLOYED column and handles the special 365243 value

APP_TRAIN_BIN_EMPLOYMENT <- application_train_slim %>%
  mutate(
    EMPLOYMENT_DURATION = case_when(
      DAYS_EMPLOYED == 365243 & NAME_INCOME_TYPE == "Pensioner" ~ "Pensioner",
      DAYS_EMPLOYED == 365243 & NAME_INCOME_TYPE == "Unemployed" ~ "Unemployed",
      DAYS_EMPLOYED > -365 ~ "Short Term",
      DAYS_EMPLOYED <= -365 & DAYS_EMPLOYED >= -1825 ~ "Medium Term",
      TRUE ~ "Long Term" # Default case if no other conditions are met
    )
  )

## Drop the rows where AMT_ANNUITY is blank because there are only a few

DROP_BLANKS <- APP_TRAIN_BIN_EMPLOYMENT %>%
  filter(!is.na(AMT_ANNUITY))


## Drop the rows where CNT_FAM_MEMBERS is blank because there are only a few

DROP_BLANKS <- DROP_BLANKS %>%
  filter(!is.na(CNT_FAM_MEMBERS))
  


## Replace NA's in OCCUPATION_TYPE with an "Unavailable"

FIX_EMPLOYMENT <- DROP_BLANKS %>%
  mutate(
    OCCUPATION_TYPE = case_when(
      is.na(OCCUPATION_TYPE) ~ "Unavailable",
      TRUE ~ OCCUPATION_TYPE
    )
  )


## Do median imputation for the missing risk score values

risk_scores <- FIX_EMPLOYMENT[, c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")]


for (col in names(risk_scores)) {
  # Convert to numeric if not already
  risk_scores[[col]] <- as.numeric(risk_scores[[col]])
  
  # Replace NA with median
  if (is.numeric(risk_scores[[col]])) {
    median_value <- median(risk_scores[[col]], na.rm = TRUE)
    risk_scores[[col]][is.na(risk_scores[[col]])] <- median_value
  }
}

FIX_EMPLOYMENT[, c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")] <- risk_scores

CLEAN_TRAIN <- clean_names(FIX_EMPLOYMENT)
view(CLEAN_TRAIN)


```






```{r}
## look at proportions of the target variable

CLEAN_TRAIN |>
     count(target) |>
     mutate(perc = n/sum(n)) 

## convert employment duration to a factor
CLEAN_TRAIN$employment_duration <- factor(CLEAN_TRAIN$employment_duration)



## set seed for reproducibility

set.seed(12345)

## split training data into a train set and a model validation set

index_numbers_split <- createDataPartition(CLEAN_TRAIN$target,p=.7,list = FALSE)

model_tr <- CLEAN_TRAIN[index_numbers_split,]
model_val <- CLEAN_TRAIN[-index_numbers_split,]



```


```{r}
# Fit logistic regression model w intercept only
(model1 <- glm(target ~ 1, 
               data = model_tr, 
               family = binomial)) %>% 
     summary() ## AIC 120383

auc_null <- roc(response = model_tr$target, predictor = predict(model1, type = "response"))
auc_null ## 0.5, totally random



# Convert target to factor with appropriate labels
model_tr$target <- factor(ifelse(model_tr$target == 0, "no_default", "default"),
                          levels = c("no_default", "default"))

model_val$target <- factor(ifelse(model_val$target == 0, "no_default", "default"),
                          levels = c("no_default", "default"))

# Fit our logistic regression model
caret_glm <- train(
  target ~ amt_income_total +
           amt_annuity +
           amt_credit +
           ext_source_1 +
           ext_source_2 +
           ext_source_3 +
           days_birth +
           employment_duration,
  method = "glm",
  family = "binomial",   # binary target
  trControl = trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,    
    sampling = "down"   # down sample the majority class
  ),
  data = model_tr,
  metric = "ROC"
)

summary(caret_glm)


# ROC for training set
train_probs <- predict(caret_glm, newdata = model_tr, type = "prob")[, "default"]
roc_train <- roc(response = model_tr$target, predictor = train_probs)
cat("Train AUC:", auc(roc_train), "\n")

# ROC for test set
test_probs <- predict(caret_glm, newdata = model_val, type = "prob")[, "default"]
roc_test <- roc(response = model_val$target, predictor = test_probs)
cat("Test AUC:", auc(roc_test), "\n")

# Optional: Plot both ROC curves
plot(roc_train, col = "blue", main = "ROC Curves")
lines(roc_test, col = "red")
legend("bottomright", legend = c("Train", "Test"), col = c("blue", "red"), lwd = 2)

```


```{r}


# Random Forest

caret_rf <- train(
  target ~ amt_income_total +
           amt_annuity +
           amt_credit +
           ext_source_1 +
           ext_source_2 +
           ext_source_3 +
           days_birth +
           employment_duration,,
  method = "ranger",   # Fast Random Fores
  trControl = trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    sampling = "down"
  ),
  data = model_tr,
  metric = "ROC"
)

summary(caret_rf)
view(caret_rf)


# ROC for training set
train_probs <- predict(caret_rf, newdata = model_tr, type = "prob")[, "default"]
roc_train <- roc(response = model_tr$target, predictor = train_probs)
cat("Train AUC:", auc(roc_train), "\n")

# ROC for test set
test_probs <- predict(caret_rf, newdata = model_val, type = "prob")[, "default"]
roc_test <- roc(response = model_val$target, predictor = test_probs)
cat("Test AUC:", auc(roc_test), "\n")

# Optional: Plot both ROC curves
plot(roc_train, col = "blue", main = "ROC Curves")
lines(roc_test, col = "red")
legend("bottomright", legend = c("Train", "Test"), col = c("blue", "red"), lwd = 2)

# ROC for training set
train_probs <- predict(caret_rf, newdata = model_tr, type = "prob")[, "default"]
roc_train <- roc(response = model_tr$target, predictor = train_probs)
cat("Train AUC:", auc(roc_train), "\n")

# ROC for test set
test_probs <- predict(caret_rf, newdata = model_val, type = "prob")[, "default"]
roc_test <- roc(response = model_val$target, predictor = test_probs)
cat("Test AUC:", auc(roc_test), "\n")

# Optional: Plot both ROC curves
plot(roc_train, col = "blue", main = "ROC Curves")
lines(roc_test, col = "red")
legend("bottomright", legend = c("Train", "Test"), col = c("blue", "red"), lwd = 2)



```
```{r}
library(gbm)

# Gradient Boosting Machine

caret_xgb <- train(
  target ~ .,
  method = "gbm",
  trControl = trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    sampling = "down"
  ),
  data = model_tr,
  metric = "ROC",
  tuneLength = 5
)

summary(caret_xgb)
view(caret_xgb)


# ROC for training set
train_probs <- predict(caret_xgb, newdata = model_tr, type = "prob")[, "default"]
roc_train <- roc(response = model_tr$target, predictor = train_probs)
cat("Train AUC:", auc(roc_train), "\n")

# ROC for test set
test_probs <- predict(caret_xgb, newdata = model_val, type = "prob")[, "default"]
roc_test <- roc(response = model_val$target, predictor = test_probs)
cat("Test AUC:", auc(roc_test), "\n")

# Optional: Plot both ROC curves
plot(roc_train, col = "blue", main = "ROC Curves")
lines(roc_test, col = "red")
legend("bottomright", legend = c("Train", "Test"), col = c("blue", "red"), lwd = 2)



# Extract variable importance
importance <- varImp(caret_xgb, scale = TRUE)

# Convert to data frame for ggplot
importance_df <- data.frame(
  Feature = rownames(importance$importance),
  Importance = importance$importance$Overall
)

# Plot of the most important predictors
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance (GBM Model)",
       x = "Features",
       y = "Importance Score") +
  theme_minimal()

```
